---
title: "final notebook"
output: html_notebook
---

The functions of the run_analyis.R script is explained here

GOAL:

Obtain and clean data from a wearables study. Only data with mean or STD will be kept. A new tidy table will be created and saved as tidy_data.txt

step one: import libraries


```{r}
## libraries needed for melt and cast
library(reshape2)
library(reshape)
```

step two: get data
```{r}
## get data
filename <- "project_data.zip"

## download data if it does not exist
if (!file.exists(filename)){
  url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
  download.file(url,filename, method = "curl")
}

## unzip if not unzipped
if (!file.exists("UCI HAR Dataset")){
  unzip(filename)
}
```

step three: get labels

```{r}
## get activity labels
activity_labels <- read.table("UCI HAR Dataset/activity_labels.txt")

## get feature labels
features <- read.table("UCI HAR Dataset/features.txt")
head(features)
```

step four: select wanted features

```{r}
## select features with mean and std (except "meanFreq")
features_wanted <-intersect(grep(".*mean.*|.*std.*", features[,2]), 
                            grep(".*meanFreq.*", features[,2], invert = TRUE))

## get names of features
features_wanted.names <-features[features_wanted,2]
## swap the "-" for "_"
features_wanted.names <- gsub('-', '_', features_wanted.names)
## remove "()"
features_wanted.names <- gsub('[()]', '', features_wanted.names)
head(features_wanted.names)
```

step five: load train data and combine it all

```{r}
## load train data
train <- read.table("UCI HAR Dataset/train/X_train.txt")[features_wanted]
train_activities <- read.table("UCI HAR Dataset/train/Y_train.txt")
train_subjects <- read.table("UCI HAR Dataset/train/subject_train.txt")
train <- cbind(train_subjects, train_activities, train)
head(train)
```

step six: load test data and combine it all

```{r}
## load test data
test <- read.table("UCI HAR Dataset/test/X_test.txt")[features_wanted]
test_activities <- read.table("UCI HAR Dataset/test/Y_test.txt")
test_subjects <- read.table("UCI HAR Dataset/test/subject_test.txt")
test <- cbind(test_subjects, test_activities, test)
head(test)
```

step seven: combine test and train and rename cols

```{r}
## merge rows
all_data <- rbind(train, test)
## rename cols
colnames(all_data) <- c("subject", "activity", features_wanted.names)
head(all_data)
```

step eight: relabel and set both activities and subjects as factors

```{r}
all_data$activity <- factor(all_data$activity, levels = activity_labels[,1], labels = activity_labels[,2])
all_data$subject <- as.factor(all_data$subject)
head(all_data)
```
step nine: melt data

```{r}
## melt the data
all_data.melted <- melt(all_data, id = c("subject", "activity"))
head(all_data.melted)
```
step ten: recast into a clean tidy data set with each column being the mean

```{r}
## cast into a tidy dataset
tidy_data <- dcast(all_data.melted, subject + activity ~ variable, mean)
head(tidy_data)
```
step eleven: save to "tiny_data.txt"

```{r}
write.table(tidy_data, "tidy_data.txt", row.names = FALSE)
```

This data is considered tidy because each row is one observation and each column is one variable. the variable names correlate to their measurements and more info can be found in the "features_info.txt" and the README within UCI HAR Dataset

